\section{Markup language}\hypertarget{markup-language}{}\label{markup-language}

A \textbf{markup language} is a system for [[annotation\textbar{}annotating]] a [[document]]
in a way that is [[Syntax (logic)\textbar{}syntactically distinguishable]] from the
text.\footnote{\{\{MerriamWebsterDictionary\textbar{}markup language\}\}} The idea and terminology
evolved from the ``marking up'' of paper manuscripts, i.e., the revision
instructions by editors, traditionally written with a [[blue pencil (editing)\textbar{}blue pencil]]
on authors' [[manuscript]]s.

In [[digital media]] this ``blue pencil instruction text'' was replaced by
[[HTML element\textbar{}tags]], that is, instructions are expressed directly by tags or
``instruction text encapsulated by tags.'' Examples include typesetting instructions
such as those found in [[troff]], [[TeX]] and [[LaTeX]], or structural markers
such as [[Extensible Markup Language\textbar{}XML]] tags. Markup instructs the software
that displays the text to carry out appropriate actions, but is omitted from
the version of the text that users see.

Some markup languages, such as the widely used [[HTML]], have pre-defined
[[presentation semantics]]-meaning that their specification prescribes how
to present the [[structured data]]. Others, such as XML, do not have
them and are general purpose.

[[Hypertext\textbar{}HyperText]] Markup Language (HTML), one of the document formats
of the [[World Wide Web]], is an instance of [[Standard Generalized Markup Language\textbar{}SGML]]
(though, strictly, it does not comply with all the rules of SGML),
and follows many of the markup conventions used in the publishing industry
in the communication of printed work between authors, editors, and printers.

\subsection{Types}\hypertarget{types}{}\label{types}

There are three main general categories of electronic markup:\footnote{\{\{cite journal\textbar{}date=November 1987\textbar{}title=Markup systems and the future of scholarly text processing\textbar{}journal=[[Communications of the ACM]]\textbar{}publisher=[[Association for Computing Machinery\textbar{}ACM]]\textbar{}volume=30\textbar{}issue=11\textbar{}pages=933–947\textbar{}doi=10.1145/32206.32209\textbar{}url=http://xml.coverpages.org/coombs.html\textbar{}first1=James H.\textbar{}last1=Coombs\textbar{}first2=Allen H. \textbar{}last2=Renear\textbar{}first3=Steven J. \textbar{}last3=DeRose\}\}}\footnote{\{\{cite web\textbar{}last=Bray\textbar{}first=Tim\textbar{}authorlink=Tim Bray\textbar{}url=http://www.tbray.org/ongoing/When/200x/2003/04/09/SemanticMarkup\#p-1\textbar{}title=On Semantics and Markup, Taxonomy of Markup\textbar{}website=www.tbray.org/ongoing\textbar{}date=9 April 2003\textbar{}access-date=9 July 2015\}\}}

Presentational markup
:The kind of markup used by traditional word-processing systems: binary codes embedded within
document text that produce the [[WYSIWYG]] effect. Such markup is usually hidden from human
users, even authors or editors.

Procedural markup
:Markup is embedded in text and provides instructions for programs that are to process the text.
Well-known examples include [[troff]], [[TeX]], and [[PostScript]]. It is expected that the
processor will run through the text from beginning to end, following the instructions as
encountered. Text with such markup is often edited with the markup visible and directly
manipulated by the author. Popular procedural-markup systems usually include programming
constructs, so macros or subroutines can be defined and invoked by name.

Descriptive markup
:Markup is used to label parts of the document rather than to provide specific instructions as to
how they should be processed. Well-known examples include [[LaTeX]], [[HTML]], and [[XML]]. The
objective is to decouple the inherent structure of the document from any particular treatment or
rendition of it. Such markup is often described as ``semantic''. An example of descriptive markup
would be HTML's \texttt{\textless{}cite\textgreater{}} tag, which is used to label a citation. Descriptive markup—sometimes
called \emph{logical markup} or \emph{conceptual markup}-encourages authors to write in a way that
describes the material conceptually, rather than visually.\footnote{Michael Downes. [http://www.ams.org/notices/200211/comm-downes.pdf ``TEX and LATEX 2e'']}

There is considerable blurring of the lines between the types of markup.
In modern word-processing systems, presentational markup is often saved in
descriptive-markup-oriented systems such as [[XML]], and then processed procedurally
by implementations. The programming constructs in procedural-markup systems such as
[[TeX]] may be used to create higher-level markup systems that are more descriptive,
such as [[LaTeX]].

In recent years, a number of small and largely unstandardized markup languages
have been developed to allow authors to create formatted text via web browsers,
for use in [[wiki]]s and web forums. These are sometimes called [[lightweight markup language]]s.
[[Markdown]] or the markup language used by [[Wikipedia]] are examples of such [[wiki markup]].

\subsection{History}\hypertarget{history}{}\label{history}

\subsubsection{Etymology and origin}\hypertarget{etymology-and-origin}{}\label{etymology-and-origin}

The term \emph{markup} is derived from the traditional publishing practice of \emph{``marking up''}
a [[manuscript]], which involves adding handwritten annotations in the form of conventional
symbolic [[Printing\textbar{}printer]]'s instructions in the margins and text of a paper manuscript or printed.
it is computer jargon used in coding [[Proofreading\textbar{}proof]]. For centuries, this task was done
primarily by skilled typographers known as ``markup men''\footnote{Allan Woods, \emph{Modern Newspaper Production} (New York: Harper \& Row, 1963), 85; Stewart Harral, \emph{Profitable Public Relations for Newspapers} (Ann Arbor: J.W. Edwards, 1957), 76; and \emph{Chiarella v. United States}, \{\{ussc\textbar{}445\textbar{}222\textbar{}1980\}\}.}
or ``copy markers''\footnote{\emph{From the Notebooks of H.J.H \& D.H.A on Composition}, Kingsport Press Inc., undated (1960s).}
who marked up text to indicate what [[typeface]], style, and size should be applied to each part, and then passed the manuscript to others for [[typesetting]] by hand. Markup was also commonly applied by editors, proofreaders, publishers, and graphic designers, and indeed by document authors.

\subsubsection{GenCode}\hypertarget{gencode}{}\label{gencode}

The first well-known public presentation of markup languages in computer text processing was made
by [[William W. Tunnicliffe]] at a conference in 1967, although he preferred to call it \emph{generic coding.}
It can be seen as a response to the emergence of programs such as [[RUNOFF]] that each used their
own control notations, often specific to the target typesetting device. In the 1970s, Tunnicliffe
led the development of a standard called GenCode for the publishing industry and later was the
first chair of the [[International Organization for Standardization]] committee that created
[[SGML]], the first standard descriptive markup language. Book designer Stanley Rice published
speculation along similar lines in 1970.\footnote{Rice, Stanley. ``Editorial Text Structures (with some relations to information structures and format controls in computerized composition).'' American National Standards Institute, March 17, 1970.}
[[Brian Reid (computer scientist)\textbar{}Brian Reid]], in his 1980 dissertation at
[[Carnegie Mellon University]], developed the theory and a working implementation
of descriptive markup in actual use.

\ldots{}
